---
title: "San Francisco Crime"
---

In this Kaggle competition the category of crime that occurred in San Francisco was predicted. 
It was my first Data Science related project during my studies, which I used to understand the steps that should be taken to solve a prediction task. 

Here it can be found a visualization part of the project.

All of the data was used from the Kaggle website.

Firstly, necessary packages for the project were imported:

```{r Loading, a_taller_figure, message = FALSE, echo = TRUE, warning=FALSE, results='hide'}

library(ggplot2)
library(vcd)
library(dplyr)
library(tidyr)
library(lubridate)
library(scales)
library(ggmap)
library(caret)
library(randomForest)
library(data.table)
library(e1071)
library(LiblineaR)
library(knitr)
library(formatR)
library(kableExtra)

```

### **Training Data preparation**

Loading data and checking the structure, the number of dimensions, names of variables and the summary of the training data:
```{r dataset_summary, a_taller_figure, message = FALSE, warning=FALSE, results='show'}
# Import data

train <- read.csv("/Users/natadi/Documents/San Francisco Crime/train.csv")

# Training dataset summary
train %>%
  head() %>%
  knitr::kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover","responsive")) 

dim(train)
names(train)

```

#### **Missing values**

```{r missing_values, a_taller_figure, message = FALSE, warning=FALSE, results='show'}
sum(is.null(train))
sum(is.na(train))
```

Splitting the column **Dates** from the training set in two columns: **Date** and **Time**.

Creating additional variables **Year**, **Month** and **Hour** to make it easier to visualize, analyse data and build prediction models.

```{r dates_proc, a_taller_figure, message = FALSE, warning=FALSE, results='show'}
# "Dates" column split
train <- separate(train, Dates, c("Date", "Time"), sep =" ", remove = TRUE)

# New variables creation
train$Year <- as.factor(year(strptime(train$Date, "%Y-%m-%d")))
train$Month <- as.factor(month(strptime(train$Date, "%Y-%m-%d")))
train$Hour <- as.factor(hour(strptime(train$Time,"%H:%M:%S")))
str(train) 

```


### **Data Visualization**

#### **Frequency of Crimes in San Francisco**

The histogram shows the most occured categories of crimes are: "Larceny/Theft", "Other Offences", "Non-Criminal", "Assault", "Vehicle Theft", "Drug/Narcotic", "Warrants", "Vandalism", "Burglary", "Suspicious Occ","Robbery" and "Fraud". 

It can be concluded that these categories appeared in around of 80% of all events.

```{r freq_hist, a_taller_figure, message = FALSE, warning=FALSE, results='show'}
ggplot(data = train,aes(x = Category)) +  
    geom_bar(fill = "steelblue3") +
  ggtitle("Frequency of Crimes in San Francisco") +
  xlab("")+
  ylab("Count") +
  coord_flip() +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 8),
    legend.position = "None",
    panel.background = element_rect(fill = "white"), 
    panel.grid.major = element_line(colour = "grey90"),
    axis.line = element_line(size = 0.4, linetype = "solid")) +
   scale_y_continuous(labels = scales::comma) 

```

#### **Number of Crimes by a weekday**

This histogram demonstrates that on days as Friday, Wednesday and Saturday the most crimes took a part:

```{r wd_freq_hist, a_taller_figure, message = FALSE, warning=FALSE, results='show'}
DOW <- count(train, DayOfWeek)
summary(train$DayOfWeek)
DOW$DayOfWeek <- factor(DOW$DayOfWeek, levels = c("Monday", "Tuesday", "Wednesday", 
                                                 "Thursday", "Friday", "Saturday", "Sunday"))

ggplot(DOW, aes(x = reorder(DayOfWeek,-n), y = n, fill = n)) +  
  geom_bar(
      stat = "identity", 
      width = 0.9,
      color = "slategray") +
  geom_text(
    aes(label= n),
    color="black", 
    vjust=1.6, 
    size=3.5) +
  scale_fill_gradient(
     low="steelblue1", high="steelblue4") +
  theme(
    legend.position = "None",
    panel.background = element_rect(fill = "white"), 
    panel.grid.major = element_line(colour = "grey90"),
    axis.line = element_line(size = 0.4, linetype = "solid"),
    axis.text.x = element_text(size=11),
    axis.text.y = element_text(size = 10)) +
   scale_y_continuous(labels = comma) +
  ggtitle("Number of crimes by a day of week") +
  xlab("")+
  ylab("Count") 
```

#### **Number of Crimes per a district**

The graph shows that the most dangerous districts are "Southern", "Mission" and "Northern".

```{r distr_freq_hist, a_taller_figure, message = FALSE, warning=FALSE, results='show'}

distr <- count(train, PdDistrict)
head(distr)
levels(train$PdDistrict)
ggplot(distr, aes(x = reorder(PdDistrict,-n), y = n, fill = n)) +  
    geom_bar(
      stat = "identity", 
      width = 0.9) +
  scale_fill_gradient(low="steelblue1", high="steelblue4") +
  theme(
    axis.text.x = element_text(
      angle = 65,
      vjust = 0.5,
      size=11),
    axis.text.y = element_text(size = 10),
    legend.position = "None",
    panel.background = element_rect(fill = "white"), 
    panel.grid.major = element_line(colour = "grey90"),
    axis.line = element_line
    (size = 0.4, linetype = "solid")) +
   scale_y_continuous(labels = scales::comma)+
   geom_text(
    aes(label= n),
    color="black", 
    vjust=1.6, 
    size=3.5) + 
  ggtitle("Number of crimes in different districts") +
  xlab("")+
  ylab("Count")

```

#### **Time graphs**

The most crimes were comitted in 2013.

```{r year_hist, a_taller_figure, message = FALSE, warning=FALSE, results='show'}
ggplot(data=train, aes(x=Year)) +
  geom_bar(colour="black", fill="steelblue2") +
  ylab("Count") +
  xlab("") +
  ggtitle("Number of Crimes in a Year") +
  theme(
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 10),
    legend.position = "None",
    panel.background = element_rect(fill = "white"), 
    panel.grid.major = element_line(colour = "grey90"),
    axis.line = element_line(size = 0.4, linetype = "solid")) +
   scale_y_continuous(labels = scales::comma)

```


Crimes tend to happen at 12 am, between 4 pm and 7 pm and in the midnight.

```{r hour_hist, a_taller_figure, message = FALSE, warning=FALSE, results='show'}
ggplot(data=train, aes(x=Hour)) +
  geom_bar(colour="black", fill="steelblue2") +
  ylab("Count")+
  xlab("") +
  ggtitle("Number of Crimes in an Hour") +
  theme(
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 10),
    legend.position = "None",
    panel.background = element_rect(fill = "white"), 
    panel.grid.major = element_line(colour = "grey90"),
    axis.line = element_line(size = 0.4, linetype = "solid")) +
   scale_y_continuous(labels = scales::comma)

```


The graph below shows that in spring months and in October crimes occured the most. 

```{r month_hist, a_taller_figure, message = FALSE, warning=FALSE, results='show'}
ggplot(data=train, aes(x=Month)) +
  geom_bar(colour="black", fill="steelblue2") +
  ylab("Count") +
  xlab("") +
  ggtitle("Number of Crimes in a Month") +
  theme(
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 10),
    legend.position = "None",
    panel.background = element_rect(fill = "white"), 
    panel.grid.major = element_line(colour = "grey90"),
    axis.line = element_line(size = 0.4, linetype = "solid")) +
   scale_y_continuous(labels = scales::comma)

```

